### 2020-02-10 paper/sex_selection_spacing-ver3.tex
New Conclusion (too long)

Consistent with a larger bias in the fertility rate when age of marriage and the length of 
birth intervals are increasing, the absolute bias is least in the first and the last 
period and highest in the middle two periods.
Hence, the fertility rate declined too fast between the mid-1980s and the end of the 
century, and only recently as the rate of increase for the birth intervals have slowed 
are the predicted cohort fertility and the fertility beginning to converge again.
Even with the convergence, the predicted cohort fertility for the last period is above 
the 1992--93 fertility rate for every group except urban women with no education.
Furthermore, for the last period, the predicted cohort fertility is stil between ten
and 20 percent higher than the fertility rate.

@@

Despite the declines in the predicted cohort fertility based on the hazard model, 
predicted cohort fertility is still mostly above the generally accepted replacement 
level of 2.1.
Only for urban women with twelve or more years of education is the predicted cohort
fertility below 2.1, and even then it is still more than 0.3 children higher than the 
fertility rate estimate of 1.5.
Furthermore, because only the first four births are used, the predicted cohort fertility
numbers are likely too low, especially for the lower education groups.

@@

[TK things to cover:
- almost everybody marry, so these numbers should reflect cohort fertility relatively well
- what happens when women begin to join the labor force in larger numbers?
- better job oppportunities may be a direct way to affect sex selection since it makes 
it more expensive to have long spacing
- Incorporate Newman 1984 discussion (p 947) on income vs substitution effect
- why has fertility fallen at all? Lower mortality, higher quality children?
- lower mortality and a higher than previously anticipated fertility implies that 
population growth in India is higher than expected.
]

[TK future research:
- health effects of multiple abortions on women and prior children
- better address selection effects for mortality estimations
- long-term outcomes for children
- labor force participation
- ]

[from fertility section]
Some tempo effects, such as those that arise from postponement of childbearing, are 
studied extensively in the literature \citep{Bongaarts1999}.
There are, however, to the best of my knowledge, no other cases where there has been
as substantial an increasing in spacing as for India.

[from mortality section]
Furthermore, with the adverse effects of shorter spacing 
strongly mitigated by higher education and no apparent 
increase in mortality risk from sex selection,
better-educated women can both try to conceive early
and and use multiple abortion in the absence of a boy
without substantial increasing mortality risk.

>>>>

Consistent with a larger bias in the fertility rate when age of marriage and the length of 
birth intervals are increasing, the absolute bias is least in the first and the last 
period and highest in the middle two periods.
Hence, the fertility rate declined too fast between the mid-1980s and the end of the 
century, and only recently, as the rate of increase for the birth intervals have slowed, 
are the predicted cohort fertility and the fertility beginning to converge again.
Even with the convergence, however, the predicted cohort fertility for 2005--16 is 
above the 1992--93 fertility rate for every group except urban women with no education.
Furthermore, for the last period, the predicted cohort fertility is stil between ten
and 20 percent higher than the fertility rate.

@@

Despite the declines in the predicted cohort fertility it is still mostly above the 
generally accepted replacement level of 2.1.
Only for urban women with twelve or more years of education is the predicted cohort
fertility below 2.1, and even then it is still more than 0.3 children higher than the 
fertility rate estimate of 1.5.
Furthermore, because only the first four births are used, the predicted cohort fertility
numbers are likely too low, especially for the lower education groups.

@@

% What has happened to spacing and why

One effect of the major changes that India has experienced over the last four decades is
that spacing between births have increased for everybody.
The increases in spacing are larger the higher the parity and the higher the mother's 
education and are larger in urban than in rural areas.

The most substantial increases in birth intervals arise, however, with heavy use of sex 
selection, as, for example, urban women with twelve or more years of education and two 
girls.
We, therefore, now in some cases observe a reversal of the traditional spacing
pattern where strong son preference was associated with shorter birth intervals the fewer 
sons a family has \citep{ben-porath76b,Leung1988}.
This reversal comes about because couples with a strong preference for sons now use sex 
selection, not because son preference has changed over time.

Despite the longer average birth intervals, a large number of women still have very
short birth intervals.
Hence, the changes in birth spacing is driven by the longer intervals becoming longer
over time, both for women who do not use sex selection, but especially for those who
do and end up with multiple abortions.
The exception to the small change in short spacing is the delay in conception for
women with very high sex selection use, presumably because this group knows that the
next child will be a boy.

% fertility
Fertility has declined for all groups, but the likelihood of having an additional child
still depends strongly on the number of sons, with women with no sons having the highest 
probabilities.

That those women most likely to use sex selection are also the most likely to have a
next birth, combined with the secular increases in spacing, makes the total fertility 
rate a more biased measure of cohort fertility.
This bias was most prominent early in the spread of sex selection, where the fertility
rate was up to one child lower than the predicted cohort fertility, but is still
present with the predicted cohort fertility ten to twenty percent higher than 
the fertility rate.

Tempo effects are studied extensively in the literature,%
\footnote{
See, for example, \citet{Bongaarts1999}.
}
but there are, however, to my knowledge, no other cases where there has been as 
substantial an increase in birth intervals and associated bias in fertility rates 
as for India.
It is, therefore, conceivable that we might see increases in TFR as birth spacing 
stabilizes or even declines again if interventions against sex selection are successful.

The only group who have a predicted cohort fertility below replacement are the 
best-educated urban women, with a predicted fertility of 1.8. 
Note that the predicted cohort fertility estimates come with the caveats that they are
based on married Hindu women and the model only includes births up to parity four.
However, almost everybody marry and Hindu constitute 80 percent of the population,
and since only the first fourth parities are used, it is possible that these 
estimates are lower-bound, especially for the less educated women.

% Mortality
There has been a substantial reduction in infant mortality over time with the size of
the reductions larger the less educated the mother.
This convergence means that there is now little difference in mortality risk across
education groups if the birth took place more than 21 months from the prior birth.
Short birth spacing is still more dangerous for the child, with the least increase for
the best-educated women.
Despite the potential that extensive use of sex selection might negatively affect the 
next child's survival, there is no evidence that children born to women who are the most
likely to use sex selection have worse outcomes. 
The rapid reduction in infant mortality risk, combined with the longer spacing,
which further reduces mortality, and higher than previously realized fertility, 
implies that population growth in India is, and will be, higher than expected.

% Future research and policy

The results here lead to two important questions that future research should address.

First, how does the changing birth spacing affect prior children?
We know that girls' health outcomes improve in the presence of sex selection 
\citep{Lin2014,Hu2015}.
An extensive use of sex selection implies is that only-girls families are much less 
likely to have a very short birth interval, which may reduce sibling competition for 
resources.
Hence, better health outcomes for girls could be an unintended side-effect 
of the longer spacing that arises from sex-selective abortions, rather than because the 
smaller number of girls makes them more valued as is often assumed.
A more positive reading would be that parents are more likely to view prior girls as
wanted since they know that they will eventually have a son.
In either case, an effective ban on sex selection could, at least temporarily, worsen
health outcomes for girls.

Second, is there an interaction between female labor force participation and 
sex selection?
Women may be staying out of the labor market exactly because the easier access to sex 
selection means that they are now more likely to have a boy and because the expected 
spacing between births has increased.
The higher return to male offspring's human capital combined with access to sex
selection may also explain another puzzle, why is the labor force participation so
low when fertility is low and education has increased.
As shown here, part of the answer may be that the number of children is, in fact, not as 
low as previously thought, both because of the increased spacing and that more 
children survive.

Better job opportunities for women may be a direct way to affect sex selection for two 
reasons.
First, it makes it more expensive to be out of the labor market for long periods.
Second, it would moderate the differential in potential earnings between husband and
wife and make it more attractive to invest in female compared to male offspring's human 
capital.
This approach may, however, be a double-edged sword.
If better job opportunities further lower desired fertility the result may be even
higher use of sex selection, everything else equal.
Hence, understanding the trade-off between long-term benefits from improvements in 
women's labor force participation and short-term costs from potential increases in sex 
selection is of paramount importance.
